---
title: "ST5005 — Section 1: Exploratory Data Analysis in R"
subtitle: "Self-study companion"
author: "Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo=TRUE, fig.height=3.5, fig.width=6,
                      fig.align="center", comment="")

# Make the Cork data available in all exercise chunks
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
n <- length(maxtp)
days <- 1:n
```

## The data

We use daily maximum air temperature (°C) recorded at Cork Airport during
January 2023, downloaded from Met Éireann (`met.ie`).

```{r show-data}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
n <- length(maxtp)
maxtp
```

**Try it:** How many observations are there? What were the minimum and maximum
temperatures recorded?

```{r ex-basics, exercise=TRUE}
# Edit and run this box freely
length(maxtp)
min(maxtp)
max(maxtp)
```

---

## 1. Frequency table

```{r freq-table}
breaks <- c(3, 5, 7, 9, 11, 13)
intervals <- cut(maxtp, breaks=breaks, right=FALSE, include.lowest=TRUE)
freq   <- table(intervals)
rel    <- freq / n
cumrel <- cumsum(rel)
data.frame(Frequency  = as.integer(freq),
           Relative   = round(rel,    3),
           Cumulative = round(cumrel, 3))
```

**Try it:** Change the class width from 2°C to 1°C by editing the `breaks`
vector below. How does the table change?

```{r ex-freqtable, exercise=TRUE}
breaks_new <- c(3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13)
intervals_new <- cut(maxtp, breaks=breaks_new, right=FALSE, include.lowest=TRUE)
table(intervals_new)
```

---

## 2. Histogram and box plot

```{r hist-box, fig.width=10}
par(mfrow=c(1,2))
hist(maxtp, breaks=c(3,5,7,9,11,13), col="steelblue", border="white",
     main="Daily max temperature, Cork Airport\nJanuary 2023",
     xlab="Temperature (°C)", ylab="Frequency")
abline(v=mean(maxtp),   col="red",    lwd=2, lty=1)
abline(v=median(maxtp), col="orange", lwd=2, lty=2)
legend("topleft", legend=c("Mean","Median"),
       col=c("red","orange"), lwd=2, lty=1:2, bty="n", cex=0.8)

boxplot(maxtp, col="steelblue", border="grey30", horizontal=TRUE,
        main="Box plot", xlab="Temperature (°C)", pch=19)
```

**Try it:** Change the number of histogram bins by editing `breaks` below.
Try `breaks=10` or `breaks=5`. Notice how the apparent shape changes.

```{r ex-hist, exercise=TRUE}
hist(maxtp, breaks=6, col="steelblue", border="white",
     main="Histogram — try changing breaks",
     xlab="Temperature (°C)")
abline(v=mean(maxtp),   col="red",    lwd=2, lty=1)
abline(v=median(maxtp), col="orange", lwd=2, lty=2)
```

---

## 3. Measures of centre and spread

```{r summaries}
cat("Mean:  ", round(mean(maxtp), 3), "°C\n")
cat("Median:", median(maxtp),         "°C\n")
cat("SD:    ", round(sd(maxtp),   3), "°C\n")
cat("IQR:   ", IQR(maxtp),            "°C\n")
quantile(maxtp, probs=c(0.25, 0.75))
```

The mean (≈ 8.61°C) is slightly below the median (9.0°C), consistent with the
cold spell in mid-January pulling the mean down.

**Try it:** `summary()` gives the five-number summary plus the mean in one
call. Run it, then try adding a large fake outlier to see how the mean shifts
but the median barely moves.

```{r ex-summaries, exercise=TRUE}
summary(maxtp)

# Add a fake extreme cold day and compare
maxtp_with_outlier <- c(maxtp, -5)
cat("\nWith outlier:\n")
cat("Mean:  ", round(mean(maxtp_with_outlier), 3), "\n")
cat("Median:", median(maxtp_with_outlier),          "\n")
```

---

## 4. Outlier check

```{r outliers}
Q1 <- quantile(maxtp, 0.25)
Q3 <- quantile(maxtp, 0.75)
iqr <- Q3 - Q1
lower_fence <- Q1 - 1.5 * iqr
upper_fence  <- Q3 + 1.5 * iqr
cat("Fences: [", round(lower_fence, 2), ",", round(upper_fence, 2), "]\n")
outliers <- maxtp[maxtp < lower_fence | maxtp > upper_fence]
cat("Outliers:", if (length(outliers) == 0) "none" else outliers, "\n")
```

**Try it:** Change the multiplier from 1.5 to 3.0 (Tukey's far-out rule).
Does any value get flagged?

```{r ex-outliers, exercise=TRUE}
Q1 <- quantile(maxtp, 0.25)
Q3 <- quantile(maxtp, 0.75)
iqr <- Q3 - Q1
multiplier <- 1.5   # try changing this to 3.0

lower <- Q1 - multiplier * iqr
upper <- Q3 + multiplier * iqr
cat("Fences: [", round(lower, 2), ",", round(upper, 2), "]\n")
maxtp[maxtp < lower | maxtp > upper]
```

---

## 5. ECDF and KDE

```{r ecdf-kde, fig.width=10}
par(mfrow=c(1,2))
plot(ecdf(maxtp), main="ECDF", xlab="Temperature (°C)",
     ylab=expression(hat(F)(x)), col="steelblue", lwd=2, pch=NA)

plot(density(maxtp, bw="SJ"), main="KDE (SJ bandwidth)",
     xlab="Temperature (°C)", ylab="Density",
     col="steelblue", lwd=2)
polygon(density(maxtp, bw="SJ"),
        col=adjustcolor("steelblue", 0.2), border=NA)
```

**Try it:** The bandwidth `bw` controls smoothness. Replace `"SJ"` with a
number (e.g. `0.5`, `1`, `3`) and see how the KDE changes.

```{r ex-kde, exercise=TRUE}
plot(density(maxtp, bw="SJ"), lwd=2, col="steelblue",
     main="KDE — try changing bw",
     xlab="Temperature (°C)", ylab="Density")
# Try: bw=0.5, bw=1, bw=3
```

---

## 6. Time series plot

Since we have 31 consecutive daily observations, a time series plot is also
informative.

```{r timeseries}
plot(1:31, maxtp, type="b", pch=19, col="steelblue",
     main="Cork Airport daily max temp, January 2023",
     xlab="Day of January", ylab="Temperature (°C)")
abline(h=mean(maxtp), col="coral", lty=2, lwd=1.5)
```

The cold snap around days 16–18 (3.2°C and 3.7°C) is immediately visible.

**Try it:** Add a smoothed trend line using `lines(lowess(...))`.

```{r ex-timeseries, exercise=TRUE}
plot(1:31, maxtp, type="b", pch=19, col="steelblue",
     main="Cork Airport — add a smooth trend",
     xlab="Day of January", ylab="Temperature (°C)")
abline(h=mean(maxtp), col="coral", lty=2)

# Add a LOWESS smooth — try changing f (span) between 0.2 and 0.8
lines(lowess(1:31, maxtp, f=0.4), col="red", lwd=2)
```
