---
title: "ST5005 — Section 4: Continuous Random Variables in R"
subtitle: "Self-study companion"
author: "Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo=TRUE, fig.height=3.5, fig.width=6,
                      fig.align="center", comment="")
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
```

## 1. The Normal distribution

### Probabilities and quantiles

```{r normal-probs}
# P(H_M > 190) where H_M ~ N(178, 7^2)
pnorm(190, mean=178, sd=7, lower.tail=FALSE)

# P(165 < H_M < 185)
pnorm(185, 178, 7) - pnorm(165, 178, 7)

# 90th percentile of male heights
qnorm(0.90, 178, 7)
```

**Try it:** Answer the Tutorial 6 questions for female heights
$H_F \sim N(165, 6^2)$.

```{r ex-normal-female, exercise=TRUE}
mu_f <- 165; sd_f <- 6

# P(H_F > 175)
pnorm(175, mu_f, sd_f, lower.tail=FALSE)

# 25th percentile (lower quartile)
qnorm(0.25, mu_f, sd_f)

# What height does a woman need to be taller than 90% of other women?
qnorm(0.90, mu_f, sd_f)
```

### Visualising Normal PDFs

```{r normal-plot, fig.width=9}
x <- seq(140, 210, length.out=300)
par(mfrow=c(1,2))

plot(x, dnorm(x, 178, 7), type="l", lwd=2, col="steelblue",
     main="Irish adult heights", xlab="Height (cm)", ylab="Density",
     xlim=c(140, 210))
lines(x, dnorm(x, 165, 6), lwd=2, col="coral")
legend("topright", legend=c("Male N(178,7²)","Female N(165,6²)"),
       col=c("steelblue","coral"), lwd=2, bty="n", cex=0.8)

qqnorm(maxtp, main="Q-Q plot: Cork Airport maxtp", pch=19, col="steelblue")
qqline(maxtp, col="coral", lwd=2)
```

**Try it:** Shade the area corresponding to $P(H_M > 190)$ under the Normal
curve.

```{r ex-normal-shade, exercise=TRUE}
mu <- 178; sigma <- 7
x  <- seq(mu - 4*sigma, mu + 4*sigma, length.out=300)

plot(x, dnorm(x, mu, sigma), type="l", lwd=2, col="steelblue",
     main="P(H_M > 190) shaded",
     xlab="Height (cm)", ylab="Density")

# Shade the tail
x_tail <- seq(190, mu + 4*sigma, length.out=100)
polygon(c(190, x_tail, mu + 4*sigma),
        c(0, dnorm(x_tail, mu, sigma), 0),
        col=adjustcolor("coral", 0.5), border=NA)
abline(v=190, lty=2, col="coral")
text(193, 0.02, round(pnorm(190, mu, sigma, lower.tail=FALSE), 4),
     col="coral", cex=0.9)
```

---

## 2. The Exponential distribution

$T \sim \mathrm{Exp}(\lambda = 1/8)$: time between customer arrivals (minutes).

```{r exponential}
lam <- 1/8
pexp(5,  rate=lam)                      # P(T <= 5)
pexp(15, rate=lam, lower.tail=FALSE)    # P(T > 15)
# Memoryless: P(T > 5 | T > 10) = P(T > 5)
pexp(5, rate=lam, lower.tail=FALSE)
```

**Try it:** Verify the memoryless property by simulation. Generate 100,000
exponential waiting times and check that conditioning on $T > 10$ does not
change the distribution of remaining time.

```{r ex-exp-memoryless, exercise=TRUE}
set.seed(42)
lam <- 1/8
T   <- rexp(100000, rate=lam)

# Unconditional P(T > 5)
cat("P(T > 5):            ", round(mean(T > 5), 4), "\n")

# Conditional P(T > 15 | T > 10) = P(remaining > 5 | already waited 10)
T_given_gt10 <- T[T > 10] - 10
cat("P(T > 5 | T > 10):   ", round(mean(T_given_gt10 > 5), 4), "\n")
```

**Try it:** Plot the Exponential PDF for different values of $\lambda$.

```{r ex-exp-plot, exercise=TRUE}
t_seq <- seq(0, 60, length.out=300)
plot(t_seq, dexp(t_seq, 1/8), type="l", lwd=2, col="steelblue",
     ylim=c(0, 0.3),
     main="Exponential PDFs for different λ",
     xlab="Time (minutes)", ylab="Density")
lines(t_seq, dexp(t_seq, 1/4),  col="coral",    lwd=2)
lines(t_seq, dexp(t_seq, 1/20), col="seagreen", lwd=2)
legend("topright",
       legend=c("λ=1/8 (mean=8)","λ=1/4 (mean=4)","λ=1/20 (mean=20)"),
       col=c("steelblue","coral","seagreen"), lwd=2, bty="n", cex=0.8)
```

---

## 3. Assessing Normality

```{r normality, fig.width=9}
par(mfrow=c(1,2))
hist(maxtp, col="steelblue", border="white", breaks=8,
     main="Cork Airport max temp", xlab="°C")
qqnorm(maxtp, pch=19, col="steelblue",
       main="Q-Q plot")
qqline(maxtp, col="coral", lwd=2)
```

```{r shapiro}
shapiro.test(maxtp)
```

$p < 0.05$: significant departure from Normality, consistent with the Q-Q
plot. Always use the formal test alongside the plot.

**Try it:** Generate a sample from a truly Normal distribution and check that
the Q-Q plot looks linear and Shapiro-Wilk gives $p > 0.05$.

```{r ex-normality, exercise=TRUE}
set.seed(42)
x_normal <- rnorm(31, mean=8.61, sd=2.31)  # same n, mean, sd as Cork data

par(mfrow=c(1,2))
hist(x_normal, col="steelblue", border="white",
     main="Simulated Normal sample", xlab="Value")
qqnorm(x_normal, pch=19, col="steelblue", main="Q-Q plot")
qqline(x_normal, col="coral", lwd=2)

shapiro.test(x_normal)
```

---

## 4. The `faithful` dataset — bimodal distribution

Old Faithful eruption durations illustrate why checking shape matters
before applying Normal-based methods.

```{r faithful-plot, fig.width=9}
par(mfrow=c(1,2))
hist(faithful$eruptions, breaks=20, col="steelblue", border="white",
     main="Old Faithful eruption durations",
     xlab="Duration (minutes)")
plot(density(faithful$eruptions, bw="SJ"), lwd=2, col="steelblue",
     main="KDE", xlab="Duration (minutes)", ylab="Density")
```

**Try it:** Run a Shapiro-Wilk test on `faithful$eruptions`. What does the
result tell you, and does it match what the plots show?

```{r ex-faithful, exercise=TRUE}
shapiro.test(faithful$eruptions)

# Also check the scatter plot of eruption duration vs waiting time
plot(faithful$waiting, faithful$eruptions,
     pch=19, col="steelblue",
     xlab="Waiting time (minutes)",
     ylab="Eruption duration (minutes)",
     main="Old Faithful: eruption vs waiting time")
```
