---
title: "ST5005 — Section 5: Populations, Sampling, and the CLT in R"
subtitle: "Self-study companion"
author: "Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo=TRUE, fig.height=3.5, fig.width=6,
                      fig.align="center", comment="")
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
mu    <- mean(maxtp)
sigma <- sd(maxtp)
```

## Introduction

The CLT is best understood by *watching* it happen. These simulations use
the Cork Airport temperature data as the population.

---

## 1. The population distribution

```{r population}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
mu    <- mean(maxtp)
sigma <- sd(maxtp)
cat("Population mean:", round(mu,    3), "°C\n")
cat("Population SD:  ", round(sigma, 3), "°C\n")
```

```{r pop-plot, fig.width=7, fig.height=3}
hist(maxtp, col="steelblue", border="white", breaks=8,
     main="Population: Cork Airport daily max temp, Jan 2023",
     xlab="Temperature (°C)")
```

This distribution is not Normal — right-skewed with a cold-spell tail.

---

## 2. Sampling distribution of $\bar{X}$

We take 10,000 samples of size $n$ (with replacement) and plot the
distribution of sample means.

```{r clt-sim, fig.width=10, fig.height=7}
set.seed(42)
B  <- 10000
ns <- c(2, 5, 15, 30)
par(mfrow=c(2,2))

for (n in ns) {
  xbars <- replicate(B, mean(sample(maxtp, n, replace=TRUE)))
  hist(xbars, breaks=40, col="steelblue", border="white", freq=FALSE,
       main=paste0("n = ", n, "   SE = ", round(sigma/sqrt(n), 2), "°C"),
       xlab=expression(bar(X)))
  curve(dnorm(x, mu, sigma/sqrt(n)), add=TRUE, col="coral", lwd=2)
}
```

As $n$ grows, $\bar{X}$ becomes approximately Normal (coral curve) even
though the population is not.

**Try it:** Repeat the simulation for a single value of $n$. Change $n$ and
observe how quickly the distribution of $\bar{X}$ approaches Normal.

```{r ex-clt, exercise=TRUE}
set.seed(42)
n  <- 5    # try 2, 10, 25, 50
B  <- 10000
mu    <- 8.610
sigma <- 2.315

xbars <- replicate(B, mean(sample(maxtp, n, replace=TRUE)))

hist(xbars, breaks=40, col="steelblue", border="white", freq=FALSE,
     main=paste0("Sampling distribution of X-bar, n = ", n),
     xlab=expression(bar(X)))
curve(dnorm(x, mu, sigma/sqrt(n)), add=TRUE, col="coral", lwd=2)
legend("topright", legend="Normal approximation",
       col="coral", lwd=2, bty="n")
```

---

## 3. Standard error shrinks with $n$

```{r se-table}
ns_all <- c(1, 2, 5, 10, 20, 31, 50, 100)
sigma  <- sd(maxtp)
data.frame(n  = ns_all,
           SE = round(sigma / sqrt(ns_all), 3))
```

**Try it:** Plot the SE as a function of $n$ and add a reference line at
SE = 0.5°C. What sample size would be needed to achieve that precision?

```{r ex-se, exercise=TRUE}
sigma <- 2.315
n_seq <- 1:200
se_seq <- sigma / sqrt(n_seq)

plot(n_seq, se_seq, type="l", lwd=2, col="steelblue",
     main="Standard error vs sample size",
     xlab="n", ylab="SE (°C)")
abline(h=0.5, col="coral", lty=2, lwd=1.5)
text(150, 0.55, "SE = 0.5°C target", col="coral", cex=0.85)

# Find the minimum n achieving SE <= 0.5
n_needed <- min(n_seq[se_seq <= 0.5])
cat("Minimum n for SE ≤ 0.5°C:", n_needed, "\n")
```

---

## 4. CLT probability calculation

Suppose we sample $n=25$ January days. What is $P(\bar{X} > 9.5)$?

```{r clt-prob}
n  <- 25
mu <- mean(maxtp)
se <- sd(maxtp) / sqrt(n)
cat("SE =", round(se, 4), "\n")
pnorm(9.5, mean=mu, sd=se, lower.tail=FALSE)
```

**Try it:** Verify the CLT probability by simulation, and explore how the
answer changes with different sample sizes.

```{r ex-clt-prob, exercise=TRUE}
set.seed(1)
mu <- 8.610; sigma <- 2.315
n  <- 25       # try 5, 10, 50

# Simulation
xbars <- replicate(100000, mean(sample(maxtp, n, replace=TRUE)))
cat("Simulated P(X-bar > 9.5):", round(mean(xbars > 9.5), 4), "\n")

# CLT approximation
cat("CLT approx P(X-bar > 9.5):",
    round(pnorm(9.5, mu, sigma/sqrt(n), lower.tail=FALSE), 4), "\n")
```

---

## 5. CLT for non-Normal data — Poisson example

The CLT applies to *any* population with finite variance — not just Normal
or near-Normal ones.

```{r clt-poisson, fig.width=9}
set.seed(42)
lambda <- 1.615   # fitted to road collision data
B      <- 10000

par(mfrow=c(1,3))
for (n in c(2, 10, 50)) {
  xbars <- replicate(B, mean(rpois(n, lambda)))
  hist(xbars, breaks=30, col="steelblue", border="white", freq=FALSE,
       main=paste0("Pois(1.615), n=", n),
       xlab=expression(bar(X)))
  curve(dnorm(x, lambda, sqrt(lambda/n)), add=TRUE, col="coral", lwd=2)
}
```

Even starting from the discrete, right-skewed Poisson distribution, the
sample mean converges rapidly to a Normal shape.

**Try it:** Repeat for the Exponential distribution, which is even more
skewed than the Poisson.

```{r ex-clt-exp, exercise=TRUE}
set.seed(42)
lam <- 1/8   # mean = 8 minutes
B   <- 10000

par(mfrow=c(1,3))
for (n in c(2, 10, 50)) {
  xbars <- replicate(B, mean(rexp(n, rate=lam)))
  hist(xbars, breaks=30, col="steelblue", border="white", freq=FALSE,
       main=paste0("Exp(1/8), n=", n),
       xlab=expression(bar(X)))
  curve(dnorm(x, 1/lam, sqrt(1/lam^2/n)), add=TRUE, col="coral", lwd=2)
}
```
