---
title: "ST5005 — Section 6: Point Estimation and Confidence Intervals in R"
subtitle: "Self-study companion"
author: "Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo=TRUE, fig.height=3.5, fig.width=6,
                      fig.align="center", comment="")
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
```

## 1. Point estimates

```{r point-estimates}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
n    <- length(maxtp)
xbar <- mean(maxtp)
s    <- sd(maxtp)
se   <- s / sqrt(n)

cat("n    =", n,              "\n")
cat("mean =", round(xbar, 4), "°C\n")
cat("SD   =", round(s,    4), "°C\n")
cat("SE   =", round(se,   4), "°C\n")
```

**Try it:** Compute the same estimates for just the first half of January
(days 1–15) and the second half (days 16–31). Do the two halves look similar?

```{r ex-estimates, exercise=TRUE}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)

first_half  <- maxtp[1:15]
second_half <- maxtp[16:31]

cat("First half  — mean:", round(mean(first_half),  2),
    "  SD:", round(sd(first_half),  2), "\n")
cat("Second half — mean:", round(mean(second_half), 2),
    "  SD:", round(sd(second_half), 2), "\n")
```

---

## 2. Confidence interval for the mean

### By hand

```{r ci-hand}
n      <- length(maxtp)
xbar   <- mean(maxtp)
s      <- sd(maxtp)
se     <- s / sqrt(n)
t_crit <- qt(0.975, df=n-1)

cat("t critical value (df=30):", round(t_crit, 4), "\n")
ci <- xbar + c(-1, 1) * t_crit * se
cat("95% CI: (", round(ci[1], 3), ",", round(ci[2], 3), ")\n")
```

### Using `t.test()`

```{r ci-ttest}
t.test(maxtp, conf.level=0.95)$conf.int
```

**Try it:** Compute 90% and 99% CIs and compare their widths.

```{r ex-ci-levels, exercise=TRUE}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)

for (level in c(0.90, 0.95, 0.99)) {
  ci <- t.test(maxtp, conf.level=level)$conf.int
  cat(level*100, "% CI: (", round(ci[1],3), ",", round(ci[2],3),
      ")  width =", round(diff(ci), 3), "\n")
}
```

---

## 3. Effect of sample size on CI width

```{r ci-sensitivity, fig.width=9}
s  <- sd(maxtp)
ns <- c(10, 20, 31, 50, 100, 200)

half_widths <- sapply(ns, function(ni) qt(0.975, df=ni-1) * s / sqrt(ni))

plot(ns, half_widths, type="b", pch=19, col="steelblue",
     main="CI half-width vs sample size (95% CI)",
     xlab="n", ylab="Half-width (°C)")
abline(h=1.0, lty=2, col="coral")
text(180, 1.05, "±1°C target", col="coral", cex=0.85)
```

**Try it:** Find the minimum sample size needed for a 95% CI with
half-width ≤ 1°C. Use the formula $n \geq (t^* s / E)^2$.

```{r ex-samplesize, exercise=TRUE}
s <- 2.315   # planning estimate of SD
E <- 1.0     # target half-width in °C

# Approximate using z (for planning purposes)
n_approx <- ceiling((qnorm(0.975) * s / E)^2)
cat("Approximate n needed:", n_approx, "\n")

# Verify: what is the actual CI half-width at this n?
actual_hw <- qt(0.975, df=n_approx-1) * s / sqrt(n_approx)
cat("Actual half-width at n =", n_approx, ":", round(actual_hw, 3), "°C\n")
```

---

## 4. CI coverage — what does 95% really mean?

```{r coverage, fig.width=8, fig.height=5}
set.seed(7)
pop_mean <- mean(maxtp)
n_sim    <- 50

ci_mat <- replicate(n_sim, {
  x <- sample(maxtp, 15, replace=TRUE)
  t.test(x, conf.level=0.95)$conf.int
})

contains <- ci_mat[1,] <= pop_mean & pop_mean <= ci_mat[2,]
cat("Intervals containing true mean:", sum(contains), "/", n_sim, "\n")

plot(NULL, xlim=c(0, n_sim+1), ylim=range(ci_mat),
     main="50 simulated 95% CIs (n=15 each)",
     xlab="Simulation index", ylab="Temperature (°C)")
abline(h=pop_mean, col="coral", lwd=2, lty=2)
for (i in seq_len(n_sim)) {
  col_i <- if (contains[i]) "steelblue" else "red"
  segments(i, ci_mat[1,i], i, ci_mat[2,i], col=col_i, lwd=1.5)
}
legend("topright", legend=c("Contains mean","Misses mean","True mean"),
       col=c("steelblue","red","coral"), lwd=2, lty=c(1,1,2),
       bty="n", cex=0.8)
```

**Try it:** Change the confidence level to 90% and the sample size to 25.
Does the number of intervals missing the true mean change as expected?

```{r ex-coverage, exercise=TRUE}
set.seed(7)
maxtp    <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
              11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
               9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
pop_mean <- mean(maxtp)
conf     <- 0.90   # try 0.95, 0.99
n_each   <- 25     # try 10, 15
n_sim    <- 100

ci_mat <- replicate(n_sim, {
  x <- sample(maxtp, n_each, replace=TRUE)
  t.test(x, conf.level=conf)$conf.int
})

contains <- ci_mat[1,] <= pop_mean & pop_mean <= ci_mat[2,]
cat(conf*100, "% CI: ", sum(contains), "/", n_sim,
    "contain the true mean\n")
cat("Expected:", conf * n_sim, "\n")
```
