---
title: "ST5005 — Section 7: Hypothesis Testing in R"
subtitle: "Self-study companion"
author: "Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo=TRUE, fig.height=3.5, fig.width=6,
                      fig.align="center", comment="")
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
```

## 1. One-sample $t$-test

Is the mean January maximum temperature at Cork Airport significantly
different from 9°C?

```{r one-sample}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
t.test(maxtp, mu=9, alternative="two.sided")
```

$p > 0.05$: no significant evidence the true mean differs from 9°C, despite
the sample mean being 8.61°C.

**Try it:** Test $H_0: \mu = 8$ instead. What changes? Try a one-sided test
($H_A: \mu > 8$) by setting `alternative="greater"`.

```{r ex-onesample, exercise=TRUE}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)

# Change mu and alternative below
t.test(maxtp, mu=8, alternative="two.sided")
```

---

## 2. Paired $t$-test

The `sleep` dataset: extra sleep (hours) for 10 patients under two drugs.

```{r paired}
head(sleep, 4)
d1 <- sleep$extra[sleep$group == 1]
d2 <- sleep$extra[sleep$group == 2]
t.test(d2, d1, paired=TRUE, alternative="greater")
```

Drug 2 produces significantly more additional sleep than Drug 1.

**Try it:** Compute and plot the paired differences directly, then verify
that testing the differences with a one-sample $t$-test gives the same result.

```{r ex-paired, exercise=TRUE}
d1   <- sleep$extra[sleep$group == 1]
d2   <- sleep$extra[sleep$group == 2]
diff <- d2 - d1

cat("Differences:", diff, "\n")
cat("Mean diff:  ", round(mean(diff), 4), "\n")

# One-sample t-test on differences — should match paired t-test above
t.test(diff, mu=0, alternative="greater")
```

---

## 3. Two-sample $t$-test

Do plant weights differ between control and treatment 1 in `PlantGrowth`?

```{r two-sample}
ctrl <- PlantGrowth$weight[PlantGrowth$group == "ctrl"]
trt1 <- PlantGrowth$weight[PlantGrowth$group == "trt1"]
cat("Control mean:", round(mean(ctrl), 3), "\n")
cat("Trt1 mean:   ", round(mean(trt1), 3), "\n")
t.test(ctrl, trt1, var.equal=FALSE)
```

```{r twosampl-plot}
boxplot(weight ~ group, data=PlantGrowth,
        col=c("steelblue","coral","seagreen"), border="grey30",
        main="Plant weights by group",
        xlab="Group", ylab="Weight (g)")
```

**Try it:** Compare control vs treatment 2 instead. Is there a significant
difference? Which comparison gives a lower $p$-value?

```{r ex-twosample, exercise=TRUE}
ctrl <- PlantGrowth$weight[PlantGrowth$group == "ctrl"]
trt2 <- PlantGrowth$weight[PlantGrowth$group == "trt2"]

cat("Control mean:", round(mean(ctrl), 3), "\n")
cat("Trt2 mean:   ", round(mean(trt2), 3), "\n")

t.test(ctrl, trt2, var.equal=FALSE)
```

---

## 4. Chi-squared test of independence

Library satisfaction by year of study (Tutorial 10 dataset).

```{r chisq}
mat <- matrix(c(42, 38, 29, 21,
                28, 32, 41, 49),
              nrow=2, byrow=TRUE,
              dimnames=list(c("Satisfied","Not satisfied"),
                            paste("Year", 1:4)))
mat
chisq.test(mat)
```

**Try it:** Inspect the expected frequencies and the residuals to see which
cells drive the significant result.

```{r ex-chisq, exercise=TRUE}
mat <- matrix(c(42, 38, 29, 21,
                28, 32, 41, 49),
              nrow=2, byrow=TRUE,
              dimnames=list(c("Satisfied","Not satisfied"),
                            paste("Year", 1:4)))
result <- chisq.test(mat)

cat("Expected frequencies:\n"); print(round(result$expected, 1))
cat("\nPearson residuals (O-E)/sqrt(E):\n")
print(round(result$residuals, 2))
```

---

## 5. One-way ANOVA

Compare all three groups in `PlantGrowth` simultaneously.

```{r anova}
fit <- aov(weight ~ group, data=PlantGrowth)
summary(fit)
TukeyHSD(fit)
```

The overall ANOVA is significant. Tukey HSD shows trt1 is significantly
lower than ctrl, and trt2 significantly higher.

**Try it:** Check the ANOVA assumptions using residual plots.

```{r ex-anova, exercise=TRUE}
fit <- aov(weight ~ group, data=PlantGrowth)
par(mfrow=c(1,2))
plot(fit, which=1, pch=19, col="steelblue")   # residuals vs fitted
plot(fit, which=2, pch=19, col="steelblue")   # Q-Q plot of residuals
```

---

## 6. Visualising the $p$-value

```{r pvalue-plot}
test_res <- t.test(maxtp, mu=9)
t_stat   <- test_res$statistic
df_val   <- test_res$parameter

t_seq <- seq(-4, 4, length.out=300)
plot(t_seq, dt(t_seq, df_val), type="l", lwd=2,
     main=paste0("t(", df_val, ") distribution — H₀: μ = 9"),
     xlab="t", ylab="Density")

t_abs  <- abs(t_stat)
t_tail <- seq(t_abs, 4, length.out=100)
polygon(c(t_tail, rev(t_tail)),
        c(dt(t_tail, df_val), rep(0, 100)),
        col=adjustcolor("coral", 0.5), border=NA)
t_tail2 <- seq(-4, -t_abs, length.out=100)
polygon(c(t_tail2, rev(t_tail2)),
        c(dt(t_tail2, df_val), rep(0, 100)),
        col=adjustcolor("coral", 0.5), border=NA)
abline(v=c(-t_abs, t_abs), lty=2, col="coral")
text(t_abs+0.1, 0.3, paste0("t = ", round(t_stat, 3)), adj=0, cex=0.85)
text(0, 0.1, paste0("p = ", round(test_res$p.value, 3)), cex=0.9)
```

**Try it:** Change $\mu_0$ to 10 and redraw the plot. Does the $p$-value
get smaller or larger? Why?

```{r ex-pvalue, exercise=TRUE}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)

mu0      <- 10    # try 9, 8, 7
test_res <- t.test(maxtp, mu=mu0)
t_stat   <- test_res$statistic
df_val   <- test_res$parameter

t_seq <- seq(-5, 5, length.out=300)
plot(t_seq, dt(t_seq, df_val), type="l", lwd=2,
     main=paste0("H₀: μ = ", mu0, "   p = ", round(test_res$p.value, 4)),
     xlab="t", ylab="Density")
t_abs <- abs(t_stat)
for (s in c(1,-1)) {
  t_r <- seq(s*t_abs, s*5, length.out=100)
  polygon(c(t_r, rev(t_r)), c(dt(t_r, df_val), rep(0,100)),
          col=adjustcolor("coral", 0.5), border=NA)
}
abline(v=c(-t_abs, t_abs), lty=2, col="coral")
```
