---
title: "ST5005 — Section 8: Simple Linear Regression in R"
subtitle: "Self-study companion"
author: "Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo=TRUE, fig.height=3.5, fig.width=6,
                      fig.align="center", comment="")
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
```

## 1. Fitting a simple linear regression

The `cars` dataset: stopping distance (ft) at various speeds (mph).

```{r fit}
fit <- lm(dist ~ speed, data=cars)
summary(fit)
```

Key output:

- **Slope** ($\hat{\beta}_1 = 3.93$): each extra mph is associated with 3.93 ft of additional stopping distance on average.
- **$R^2 = 0.651$**: speed explains 65.1% of the variation in stopping distance.

**Try it:** Extract individual components of the model fit.

```{r ex-fit, exercise=TRUE}
fit <- lm(dist ~ speed, data=cars)

cat("Intercept:", round(coef(fit)[1], 4), "\n")
cat("Slope:    ", round(coef(fit)[2], 4), "\n")
cat("R-squared:", round(summary(fit)$r.squared, 4), "\n")
cat("Residual SE:", round(summary(fit)$sigma, 4), "\n")
```

---

## 2. Scatter plot with fitted line

```{r scatter}
plot(dist ~ speed, data=cars, pch=19, col="steelblue",
     main="Stopping distance vs speed",
     xlab="Speed (mph)", ylab="Stopping distance (ft)")
abline(fit, col="coral", lwd=2)
legend("topleft",
       legend=c("Observed",
                sprintf("Fitted: %.2f + %.2f × speed",
                        coef(fit)[1], coef(fit)[2])),
       pch=c(19, NA), lty=c(NA, 1),
       col=c("steelblue","coral"), bty="n", cex=0.8)
```

**Try it:** Make a prediction for a car travelling at 22 mph, both manually
and using `predict()`. Add the point to the plot.

```{r ex-predict, exercise=TRUE}
fit <- lm(dist ~ speed, data=cars)

# Manual prediction
b0 <- coef(fit)[1]; b1 <- coef(fit)[2]
speed_new <- 22
cat("Manual prediction:", round(b0 + b1 * speed_new, 2), "ft\n")

# Using predict()
cat("predict() result: ",
    round(predict(fit, newdata=data.frame(speed=speed_new)), 2), "ft\n")

# Plot with prediction
plot(dist ~ speed, data=cars, pch=19, col="steelblue",
     main="Prediction at speed = 22 mph")
abline(fit, col="coral", lwd=2)
points(speed_new, predict(fit, newdata=data.frame(speed=speed_new)),
       pch=19, col="red", cex=1.5)
```

---

## 3. Confidence and prediction intervals

```{r intervals, fig.width=7}
new_speed <- data.frame(speed=seq(4, 25, by=0.5))
ci <- predict(fit, newdata=new_speed, interval="confidence")
pi <- predict(fit, newdata=new_speed, interval="prediction")

plot(dist ~ speed, data=cars, pch=19, col="steelblue",
     main="95% confidence and prediction intervals")
lines(new_speed$speed, ci[,"fit"], col="coral",    lwd=2)
lines(new_speed$speed, ci[,"lwr"], col="coral",    lwd=1, lty=2)
lines(new_speed$speed, ci[,"upr"], col="coral",    lwd=1, lty=2)
lines(new_speed$speed, pi[,"lwr"], col="seagreen", lwd=1, lty=3)
lines(new_speed$speed, pi[,"upr"], col="seagreen", lwd=1, lty=3)
legend("topleft",
       legend=c("Fitted","95% CI for mean","95% PI for new obs"),
       col=c("coral","coral","seagreen"), lwd=2, lty=c(1,2,3),
       bty="n", cex=0.8)
```

**Try it:** At speed = 15 mph, compute both intervals manually and compare
to the plot.

```{r ex-intervals, exercise=TRUE}
fit <- lm(dist ~ speed, data=cars)
new <- data.frame(speed=15)

ci <- predict(fit, newdata=new, interval="confidence", level=0.95)
pi <- predict(fit, newdata=new, interval="prediction", level=0.95)

cat("At speed = 15 mph:\n")
cat("Fitted value:  ", round(ci[,"fit"], 2), "ft\n")
cat("95% CI:  (", round(ci[,"lwr"],2), ",", round(ci[,"upr"],2), ")\n")
cat("95% PI:  (", round(pi[,"lwr"],2), ",", round(pi[,"upr"],2), ")\n")
cat("PI is wider by:", round(diff(pi[,2:3]) - diff(ci[,2:3]), 2), "ft\n")
```

---

## 4. Diagnostic plots

```{r diagnostics, fig.width=10}
par(mfrow=c(1,3))
plot(fit, which=1, pch=19, col="steelblue")  # residuals vs fitted
plot(fit, which=2, pch=19, col="steelblue")  # Q-Q plot
plot(fit, which=3, pch=19, col="steelblue")  # scale-location
```

A slight fan shape in the residuals vs fitted plot suggests mild
heteroscedasticity — variance increases with speed.

**Try it:** Fit a model with `sqrt(dist)` as the response and re-examine
the diagnostic plots. Does the heteroscedasticity improve?

```{r ex-diagnostics, exercise=TRUE}
fit_sqrt <- lm(sqrt(dist) ~ speed, data=cars)
par(mfrow=c(1,2))
plot(fit_sqrt, which=1, pch=19, col="steelblue",
     main="sqrt(dist) ~ speed: residuals vs fitted")
plot(fit_sqrt, which=2, pch=19, col="steelblue",
     main="Q-Q plot")
```

---

## 5. Is there a linear trend across January?

```{r cork-regression}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
day <- 1:31
fit_cork <- lm(maxtp ~ day)
summary(fit_cork)
```

The slope is close to zero and $p > 0.05$: no significant linear trend.

**Try it:** Plot the data with the fitted trend line, and add a LOWESS
smooth to see the non-linear pattern.

```{r ex-cork-regression, exercise=TRUE}
maxtp <- c(8.2, 8.4, 11.9, 11.6, 11.0, 10.1, 8.3,  6.4,  6.7, 11.7,
           11.1,11.1,  9.9,  9.9,  7.1,  3.7,  3.2,  5.1,  5.5,  7.5,
            9.0, 9.4,  8.5,  8.9, 10.4,  7.1,  6.1,  9.7, 10.9,  9.0, 9.5)
day <- 1:31
fit_cork <- lm(maxtp ~ day)

plot(day, maxtp, type="b", pch=19, col="steelblue",
     main="Cork Airport: daily max temp, January 2023",
     xlab="Day of January", ylab="Temperature (°C)")
abline(fit_cork, col="coral", lwd=2, lty=2)
lines(lowess(day, maxtp, f=0.4), col="red", lwd=2)
legend("topright",
       legend=c("Data","OLS trend","LOWESS smooth"),
       col=c("steelblue","coral","red"),
       lty=c(1,2,1), pch=c(19,NA,NA), bty="n", cex=0.8)
```
